---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
library(reshape2)
library(pracma)
```

# MortalityTables

<!-- badges: start -->
<!-- badges: end -->

The goal of MortalityTables is to provide generic base classes and functions to
handle all kinds of actuarial actuarial mortality tables (period and cohort life tables).
Cohort and static life tables are implemented, observed data can be used, and
existing life tables can be blended or extrapolated to derive new tables.

Furthermore, plotting functions are provided for reports and publications.

## Installation

You can install the development version of MortalityTables from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kainhofer/MortalityTables")
```

## About the package

The MortalityTables package provides the `mortalityTable` base class and
some derived classes to handle different types of mortality tables (also 
called life tables), mainly
used for life insurance. Additionally it provides a plot function to compare
multiple life tables either directly using the absolute mortalities in
log-linear plots or using relative mortalities as percentages of a given
reference table.

### Types of Life Tables

Provided types of mortality tables are:
    
* Base class
: Class `mortalityTable`
* Period life table
: Class `mortalityTable.period(ages, deathProbs, ..., baseYear=2000)`
: Death probabilities observed / predicted for one observation year;
No dependency on the bith year is assumed.
* Cohort life table using age-specific trends
: Class `mortalityTable.trendProjection`
: Death probabilities of a given base year are projected into the future
using age-specific trends $\lambda_x$. The death probability of an $x$-year old in year
`baseYear + n` is calculated as:
    $$q_x^{(baseYear+n)} = q_x^{(baseYear)} \cdot e^{-n\cdot\lambda_x}$$
    : Consequently, the death probabilities for a person born in year `YOB` can be calculated as
$$q_x^{YOB} = q_x^{(base)} \cdot e^{-(YOB+x-baseYear)\cdot \lambda_x}$$
    * Cohort life table approximation using age shift
: Class `mortalityTable.ageShift`
: Death probabilities for cohort $YOB$ are obtained by using death probabilities
for cohort $X$ and modifying the technical age with a birth-year dependent shift:
    $$q_x^{YOB} = q_{x+shift(YOB)}^{(base)}$$
    <!-- * Observed life table -->
    <!--     : Class `mortalityTable.observed` -->
    <!--     : Death probabilities observed during several years. The probabilities are -->
    <!--       stored as a matrix with observation year and age as dimensions. -->
    * Mixed life table
: Class `mortalityTable.mixed`
: Arithmetic mean of two life tables with given weights. This approach is
often used to generate unisex life tables by mixing male and female
mortalities with given weights (e.g. 70:30 or 40:60)
* Cohort life table using age-specific improvement factors
: Class `mortalityTable.improvementFactors`
: Project base life table using age-specific improvement factors.
* Pension table
: Class `pensionTable`
: Four states: active, early retirement / invalidity, old-age pension, death (with optional widow)
: All slots describe the corresponding transition probabilities by a 
: `mortalityTable`-derived object.

## Loading the MortalityTables package
```{R loading}
library("MortalityTables")
```

## Provided Data Sets

The package provides several real-life life tables published by census bureaus 
and actuarial associations around the world. You can use the function 
`mortalityTables.list` to list all available datasets (if no argument is given)
or all datasets that match the given pattern (wildcard character is *). You can 
then use `mortalityTables.load` to load either one single data set or all 
datasets that match the pattern.

```{r ListLoadTables}
# list all datasets for Austria
mortalityTables.list("Austria_*")

# Load the German annuity table DAV 2004-R
mortalityTables.load("Germany_Annuities_DAV2004R")

# Load all Austrian data sets
mortalityTables.load("Austria_*")
```

## Cohort and Period Mortality Data

Cohort mortality vectors (for a given birth year) or period death probabilities (for a given observation year) can be extracted with the functions `periodDeathProbabilities()` and `deathProbabilities()`:
    
```{r deathProbabilities}
mortalityTables.load("Austria_Annuities")
deathProbabilities(AVOe2005R.male, YOB = 1977, ages = 35:50)
deathProbabilities(AVOe2005R.male, YOB = 2023, ages = 35:50)

periodDeathProbabilities(AVOe2005R.male, Period = 2023, ages = 35:50)
```

If the mortality table is a cohort table, the trend is used to calculate the death probabilities for the given cohort or calendar year. If the table is a static life table, the period and cohort life tables will be identical. If the table is an observed table (i.e. observed death probabilities for each age and year), the data is extracted from the matrix' rows/columns or diagonals. In all cases, the user does not have use different methods for different underlying tables. 

## Plotting and Comparing Mortality Data

There are two plotting functions using ggplot: `plotMortalityTables()` and `plotMortalityTableComparisons()` to plot the absolute and relative mortalities. For absolute mortalities, the `q(x)` axis employs a log10-scale. The returned plot is a normal ggplot2 object, so all features provided by ggplot2 can be adde to the plots.

```{r Plotting, fig.width=8.5, fig.height=5}
mortalityTables.load("Austria_Annuities")
plotMortalityTables(AVOe2005R.male, AVOe2005R.male.unloaded, AVOe1996R.male, EROM.G1950.male, 
                    YOB = 1977, ages = 0:99, legend.position = c(0.5, 0.65))
plotMortalityTableComparisons(AVOe2005R.male, AVOe2005R.male.unloaded, AVOe1996R.male, EROM.G1950.male, 
                              YOB = 1977, ages = 0:99, legend.position = c(0.5, 0.65))
plotMortalityTrend(AVOe2005R.male, AVOe2005R.male.unloaded, AVOe1996R.male, AVOe1996R.male, EROM.G1950.male)
```

## Further information
For further information on how to use the package, see the ["Using the MortalityTables Package"](https://cran.r-project.org/package=MortalityTables/vignettes/using-the-mortalityTables-package.html) vignette.

